---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

### Load the data

```{r, echo=TRUE}
DF <- read.csv(unz("activity.zip", "activity.csv"))
```

### Process/transform the data

**Intervals**

The column `interval` represents 5-minute intervals.

```{r, echo=TRUE}
unique(DF$interval)
```

The intervals were coded by appending the hour and minute and then casting to integers. These can be converted to strings, padded with zeros on the left.

```{r, echo=TRUE}
DF$interval <- formatC(DF$interval, width=4, flag=0)
```

**Dates**

The column `date` represents dates in YYYY-MM-HH format. These can be converted to `Date` objects.

```{r, echo=TRUE}
DF$date <- as.Date(DF$date)
```


## What is mean total number of steps taken per day?

Ignore missing values for this section.

### Total steps per day

```{r, echo=TRUE}
daily_totals <- aggregate(steps ~ date, DF, sum)
daily_totals
```

### Histogram of total steps per day

```{r, echo=TRUE}
hist(daily_totals$steps, xlab="Total Steps", ylab="# Days",
     main="Histogram of Total Steps per Day")
```

### Mean and median steps per day

```{r, echo=TRUE}
mean(daily_totals$steps, na.rm=TRUE)
median(daily_totals$steps, na.rm=TRUE)
```


## What is the average daily activity pattern?

Ignore missing values for this section too.

### Time series plot of average steps per interval

```{r, echo=TRUE}
interval_avgs <- aggregate(steps ~ interval, DF, mean, na.rm=TRUE)
plot(strptime(interval_avgs$interval, format="%H%M"), interval_avgs$steps, type="l",
     xlab="Time", ylab="Average Steps in 5-Minute Interval", main="Time Series Plot of Daily Pattern")
```

### Interval with maximum average steps

```{r, echo=TRUE}
interval_avgs$interval[which.max(interval_avgs$steps)]
```


## Imputing missing values

### Total missing values

```{r, echo=TRUE}
sum(is.na(DF$steps))
```

### Imputation strategy

```{r, echo=TRUE}
tapply(DF$steps, DF$date, function(x) sum(is.na(x)))
```

There are 8 days completely missing, and the other days have no missing data.

Impute using the mean for each 5-minute interval.

### Apply imputation

```{r, echo=TRUE}
DF_fill <- DF
avg_steps <- rep(interval_avgs$steps, length=dim(DF)[1])
DF_fill$steps <- ifelse(is.na(DF$steps), avg_steps, DF$steps)
head(DF_fill)
```

### Total steps per day, after imputation

```{r, echo=TRUE}
daily_totals_fill <- aggregate(steps ~ date, DF_fill, sum)
hist(daily_totals_fill$steps, xlab="Total Steps", ylab="# Days",
     main="Histogram of Total Steps per Day - Imputed")
mean(daily_totals_fill$steps)
median(daily_totals_fill$steps)
```

The mean total daily steps did not change due to imputation. The median increased slightly.

The total daily number of steps only changed for the 8 days with missing data. These days were completely missing, so they were replaced with the average at each interval. Consequently, their totals are the average total.


## Are there differences in activity patterns between weekdays and weekends?

Use the imputed data for this section.

### Create a weekday/weekend factor

```{r, echo=TRUE}
DF_fill$weekday <- as.factor(ifelse(weekdays(DF_fill$date) %in% c("Saturday", "Sunday"), "weekend", "weekday"))
head(DF_fill)
```

### Time series plot of average steps per interval, panel by weekday/weekend

```{r, echo=TRUE}
interval_weekday_avgs <- aggregate(steps~interval*weekday, DF_fill, mean)
interval_weekday_avgs$interval <- strptime(interval_weekday_avgs$interval, format="%H%M", tz="UTC")

library(ggplot2)
library(scales)
ggplot(interval_weekday_avgs, aes(interval, steps)) + geom_line() + facet_grid(weekday~.) +
  scale_x_datetime(labels = date_format("%H:%M")) + theme_bw() +
  labs(title="Daily Pattern - Imputed, Weekday vs Weekend", x="Time", y="Average Steps in 5-Minute Interval")
```

The mid-morning peak is higher on weekdays. The other peaks are lower on weekdays. Activity starts a few hours earlier on weekdays and ends a few hours later on weekends.

